Local Type Inference for Polarised System F with Existentials

### 2024-04-04 oopsla24/tex/02-overview.tex
WIP: algorithm overview and the relation to SystemF

The standard \systemf types and terms must be polarized to be used in \fexists, 
we present the embedding formally in \cref{fig:polarization}.
For instance, the following types and terms of \systemf 
can be polarized in the given way:

@@

>>>>

In general, any term and type of \systemf can be embedded into \fexists.
This embedding---polarization---will be covered in more details in \cref{sec:rel-to-systemf}.
Now, we discuss a number of examples of standard \systemf terms and types
and their polarization in \fexists.

@@

The inference algorithm for \fexists is \emph{local}: it has a limited scope of
the inference information and does not propagate it between far-apart branches
of the syntax tree. Nevertheless, many difficulties appear in the inference of
polymorphic and existential types. Next, we discuss the most challenging of
these difficulties and how they are addressed in the algorithm.

\paragraph{Subtyping} One of the hardest parts of the inference is subtyping,
because this is where the polymorphic instantiation happens. In particular, it
is the subtyping relation that allows us to use the polymorphic terms in the
context where their concrete instantiations are expected: 
$[[Γ ⊢ ∀α⁺.↑α⁺ ≤ ↑Int]]$.

We are able to concentrate most of the difficulty in the subtyping relation 
and then use the polarization to make it decidable.

\paragraph{Impredicativity}
Impredicativity is another difficulty. Even with forall, the natural subtyping is undecidable. We have to find the restrictions that make it decidable. Unification allows us to solve equality constraints, and the LUB has to be solved with anti-unification. We have to design the system carefully to maintain decidability.

\paragraph{Least Upper Bound}
Now the LUB (Least Upper Bound) is not necessarily the syntactic equality. Examples, types don't have a LUB. How do we decide? The answer is anti-unification, which we explain at a high level. Now we have both forall and exists, and mixing unification and anti-unification.

### 2024-04-04 oopsla24/tex/03-declarative-system.tex
wip
WIP: algorithm overview and the relation to SystemF

\ilyam{Other properties?}

>>>>

\subsection{Relation to \systemf}
\label{sec:rel-to-systemf}

To establish a correspondence between \fexists and standard unpolarized \systemf,
we present a translation in both ways: the polarization from \systemf to \fexists
and the depolarization from \fexists to \systemf.

\paragraph{Type-level Translation}

The type depolarization (\cref{fig:depolarization}) simply forgets the polarization structure of types:
it removes the shift operators and the polarities of the free type variables.

The type polarization (\cref{fig:polarization}) is more involved, as there are multiple
ways to polarize a type: for instance, a variable $[[α]]$ can be polarized to $[[α⁺]]$ or $[[α⁻]]$.
The choice of polarization affects the execution strategy of the program.
We chose the positive polarization: every \systemf type is translated to a \emph{positive} type of \fexists.
This choice makes it smoother to lift the translation to the term level, which will be disused next.

\begin{figure}[ht]
  \begin{minipage}[t]{0.65\linewidth}
    \centering
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization}
    \label{fig:depolarization}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[t]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}
\end{figure}

\paragraph{Term-level Translation}

The term-level translations between systems are more complex.
They are defined as \emph{elaborations} of the typing derivations.
In other words, the input of the translation is not a term, 
but a whole typing derivation of this term in the corresponding system.
For brevity, we omit the formal definition of the term-level translation,
referring the reader to the appendix (\cref{sec:term-translation}).

We prove the soundness of the translation (\cref{lemma:translation-soundness}),
which in particular guarantees that the translation preserves typing.

\begin{property}[Soundness of the Translation]
    \label{lemma:translation-soundness}
    \begin{align*}
      & [[Γ ⊢ iN ≤ iM ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iN| → |iM| ]], \\
      & [[Γ ⊢ iP ≥ iQ ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iQ| → |iP| ]], \\
      & [[Γ; Φ ⊢ v : iP ⤳ ft]]  \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iP|]], \\
      & [[Γ; Φ ⊢ c : iN ⤳ ft]] \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iN|]], \\
      & [[fΓ ; fΦ ⊢ ft : fT ⤳ c]]  \text{ implies }  [[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]].
    \end{align*}
\end{property}

% \begin{figure}[h]
%   \begin{equation*}
%   \begin{aligned}
%     [[|↓iN|]] &= [[|iN|]] \\
%     [[|↑iP|]] &= [[|iP|]] \\
%   \end{aligned}
%   \qquad
%   \begin{aligned}
%     [[|α±|]] &= [[a]] \\
%     [[|iP → iN|]] &= [[|iP| → |iN|]] \\
%   \end{aligned}
%   \qquad
%   \begin{aligned}
%     [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
%     [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
%   \end{aligned}
%   \end{equation*}
%   \caption{Type depolarization}
%   \label{fig:depolarization}
% \end{figure}

% The translation from \systemf to \fexists can be done in two canonical ways
% depending on the desired polarity. The polarity will also determine the 
% execution strategy \ilyam{how?}.
% Moreover, the translation procedure depends on the \emph{polarity} of the
% free type variables. Thus, the translation functions 
% $[[-| fT |-^Γ]]$ and $[[+| fT |+^Γ]]$ are indexed by a context $[[Γ]]$, 
% containing either $[[α⁺]]$ or $[[α⁻]]$ for each variable $[[α]]$ free in $[[fT]]$.

% \begin{figure}[h]
%   \begin{equation*}
%   \begin{aligned}
%     [[-| a |-^Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
%     [[-| a |-^Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
%     [[-| fA → fB |-^Γ]] &= [[ +|fA|+^Γ → -|fB|-^Γ  ]] \\
%     [[-| ∀α.fT |-^Γ]] &= [[ ∀α⁺.-|fT|-^Γ,α⁺ ]] \\
%     [[-| ∃α.fT |-^Γ]] &= [[ ↑ ∃α⁻.+|fT|+^Γ,α⁻ ]] \\
%   \end{aligned}
%   \qquad
%   \begin{aligned}
%     [[+| a |+^Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
%     [[+| a |+^Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
%     [[+| fA → fB |+^Γ]] &= [[ ↓ +|fA|+^Γ → -|fB|-^Γ  ]] \\
%     [[+| ∀α.fT |+^Γ]] &= [[ ↓∀α⁺.-|fT|-^Γ,α⁺ ]] \\
%     [[+| ∃α.fT |+^Γ]] &= [[ ∃α⁻.+|fT|+^Γ,α⁻ ]] \\
%   \end{aligned}
%   \end{equation*}
%   \caption{Type polarization}
%   \label{fig:polarization}
% \end{figure}

### 2024-04-04 oopsla24/tex/04-algorithm.tex
WIP: algorithm overview and the relation to SystemF

  Rule \ruleref{\ottdruleATAppLetAnnLabel}
  infers a type of \emph{annotated} applicative let binder.
  First, it infers the type of the head of the application,
  ensuring that it is a \emph{thunked computation} $[[↓iM]]$.
  After that, it makes a recursive call
  to the application inference procedure,
  returning an algorithmic type $[[iM']]$, 
  that must be a subtype of the annotation $[[↑uP]]$.

  Then premise $[[Γ; Θ ⊨ iM' ≤ ↑iP ⫤ SC2]]$
  together with $[[Θ ⊢ SC1 & SC2 = SC]]$
  check that $[[iM']]$ can be instantiated to the annotated type $[[↑iP]]$,
  and if it is, the algorithm infers the type of the body in the extended context,
  and returns it as the result.

  Rule \ruleref{\ottdruleATAppLetLabel}
  works similarly to \ruleref{\ottdruleATAppLetAnnLabel}.
  However, since no annotation is given,
  the algorithm must ensure that the inferred $[[uQ]]$
  has the `canonical' minimal instantiation.
  To find it, it makes a call to the minimal instantiation algorithm 
  (\cref{sec:minimal-instantiation})
  that finds the substitution that satisfies the inferred constraints $[[SC]]$ and
  instantiates $[[uQ]]$ to the minimal (among other such instantiations)
  type $[[ [uσ]uQ ]]$.

>>>>

  Rule \ruleref{\ottdruleATAppLetAnnLabel}
  infers a type of \emph{annotated} applicative let binder.
  First, it infers the type of the head of the application,
  ensuring that it is a \emph{thunked computation} $[[↓iM]]$.
  After that, it makes a recursive call
  to the application inference procedure,
  returning an algorithmic type $[[uM']]$, 
  that must be a subtype of the annotation $[[↑iP]]$.

  Then premise $[[Γ; Θ ⊨ uM' ≤ ↑iP ⫤ SC2]]$
  together with $[[Θ ⊢ SC1 & SC2 = SC]]$
  check that $[[uM']]$ can be instantiated to the annotated type $[[↑iP]]$,
  and if it is, the algorithm infers the type of the body in the extended context,
  and returns it as the result.

  Rule \ruleref{\ottdruleATAppLetLabel}
  works similarly to \ruleref{\ottdruleATAppLetAnnLabel}.
  However, since no annotation is given,
  the algorithm must ensure that the inferred $[[uQ]]$
  has the `canonical' minimal instantiation.
  To find it, it makes a call to the minimal instantiation algorithm 
  (\cref{sec:constraint-singularity})
  that finds the substitution that satisfies the inferred constraints $[[SC]]$ and
  instantiates $[[uQ]]$ to the minimal (among other such instantiations)
  type $[[ [uσ]uQ ]]$.
